Adaptative Computation:\\ An Essay in Memory of John H. Holland

### 2015-10-27 paper.tex
tightened up 'evolutionary dynamics' and merged 'internal models with classifier systems

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic work
on integrating genetics with Darwinian selection \cite{Fisher1930}.
As Holland further read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain changing environments and use it in
order to improve their performance with respect to those environments.
Moreover, Holland recognized that in systems with {\em adaptive
  agents} of this kind, there is never a state of {\emph equilibrium}
or a final {\emph optimum} configuration.  Due to the environment's
``perpetual novelty'' (as Holland termed it), adaptation continues
forever in an open-ended way.  Holland's focus on open-ended,
non-equilibrium dynamics was in stark contrast with the mainstream
approach (at the time) in all these fields---the belief that ``solving
for'' stable equilibrium dynamics was the scientific goal.  Holland's
contrary view was that a system in stable equilibrium is essentially
{\emph dead}.

Underlying Holland's theory of adaptation are three core ideas:

\begin{itemize}
\item {\bf Populations:} Adaptation occurs in populations of
  individuals evolving (or learning) over time, in which statistics 
  can be leveraged to direct population dynamics.  (More on this below.)

\item {\bf Building Blocks:} In a population undergoing adaptation,
  individuals can be decomposed into {\emph building blocks}---sets of
  traits that are the evolutionary ``atoms'' of an individual's
  fitness or performance. As an example from biology, Holland gives
  the {\emph Krebs cycle}, a core cellular metabolic
  pathway. Similarly, in neuroscience, a building block of visual
  perception might be a group of neurons that responds to an oriented
  edge; in computer science, a building block of a sorting program
  might be a ``swap'' operator.

\item {\bf Exploitation Versus Exploration:} Successful adaptation
  requires the right tradeoff between {\emph exploitation}, in which
  tried-and-true building blocks propagate in a population, and {\emph
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  A system that focuses too much on exploitation
  risks never finding better individuals or not being able to adapt to
  changing environments.  Conversely, too much focus on exploration
  means that the system won't leverage the successful building blocks
  that have already been discovered.  There is some optimal balance
  between these two modes of processing.
\end{itemize}

@@

In particular, inspired by Bellman \cite{Bellman1961} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme {\emph
  exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme {\emph
  exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?

In \cite{Holland1973} and \cite{Holland1975}, Holland derived
an equation for such a strategy.  Let $A$ denote the arm currently
observed to have the higher payoff probability, and $B$ denote the
other arm.  Holland showed that the optimal strategy (the one that
yields maximal payoff, or minimal loss) is for the number of trials
allocated to $A$ to grow slightly faster than an exponential function
of the number of trials allocated to $B$.  Holland then showed that this
result extends naturally to the {\emph multi-armed bandit} case.

In an population undergoing adaptation, the building blocks making up
individuals can be compared to the arms of a multi-armed bandit.  Each
building block can be said to have a probability of ``payoff'' (i.e.,
contribution to a given individual's fitness).  Evaluating an
individual in an environment is like ``pulling the arms'' on a
multi-armed bandit, where the ``arms'' correspond to each of the
building blocks making up that individual.  Thus, in the process of
assigning a fitness to each individual in a population, adaptive
evolution can be seen as implicitly sampling the many building blocks
making up that collection of individuals.  The average fitness over
many individuals containing a particular building block gives an
estimate of that building block's payoff probability.

The question of how to balance exploitation and exploration---how to
optimally allocate trials to different arms based on their observed
payoff---now becomes the question of how to optimally sample in the
vast space of possible building blocks, based on their observed
average fitness.  Of course adaptive evolution deals in populations of
individuals, not building blocks.  There is no explicit mechanism for
keeping statistics on building blocks' ``observed average fitness''.
However, Holland's central idea here is that an approximation to
optimal building-block sampling does indeed occur, as an emergent
property of the population dynamics.

To show this mathematically, Holland defined an idealization of
adaptive evolution: an algorithm he called a {\emph reproductive
  plan}.  A reproductive plan is a population-based stochastic process
operating on bit strings, involving the ``genetic'' operators of
reproduction, fitness-based selection, crossover, mutation, and
inversion.  Building blocks are formalized as {\emph
  schemata}---patterns of bits within a string. (Here are two examples
of schemas: ``all strings beginning with the bits 10'' or ``all
strings that start with the pattern 1*1'', where * can be replaced by
either 0 or 1).

Holland's most important result was the following: He proved
mathematically that his reproductive plan results---implicitly---in a
near-optimal allocation of trials to schemata, thus optimizing the
exploitation versus exploration balance.  Holland noted that the
explicit act of assigning fitnesses to the individuals in a population
was actually implicitly sampling a much larger collection of building
blocks.  Holland termed this {\emph implicit parallelism}; the
parallel leveraging of statistics from implicit sampling is a main
strength of his population-based approaches to search.

It was this formal setting and resulting theorems that led to the
invention of genetic algorithms, which featured stochastic
population-based search and {\emph crossover} as a critical
operation: Crossover recombined successful building blocks and
thus allowed them to be tested in new contexts.  What made GAs unique
among other evolution-inspired algorithms at the time were the
mathematical foundations described above, including bit-string
representations, the focus on recombination as a central mechanism,
and a focus (at least in Holland's mind) on continual adaptation to
non-stationary environments rather than optimization to static
environments.

It is important to point out again that the framework Holland
developed in \cite{Holland1975} was more general than the genetic algorithm;
it was created in order to develop an interdisciplinary theory of
adaptation, one that would inform biology, say, as much as computer
science \cite{Christiansen1998}.  The later, successful
application of genetic algorithms to real-world optimization and
learning tasks was, for Holland, just icing on the cake.

\section{Learning Via Credit Assignment and Rule Discovery}

While genetic algorithms are nowadays most often associated with models of
biological evolution, in Holland's view the processes of adaptation he
was studying manifested themselves in many different complex systems.
One system of special interest to Holland was the human mind, and in
particular, its capacity to learn via induction.

In the 1970s and 80s Holland formulated and extensively explored a
particular model of decision-making, action, and inductive learning,
called the {\emph classifier system} (e.g.,
\cite{Holland1977,Holland1986}.  In a classifier system, a population
of ``if-then'' rules interacts with an environment, and over time is
able to learn to improve its performance via both {\emph
  credit-assignment} and {\emph rule-discovery} algorithms.

Like genetic algorithms, classifier systems can be viewed both as 
models of adaptation and as artificial-intelligence (AI) methods.  At
the time Holland was developing classifier systems, the field of AI
was focused on expert systems, which typically did not learn on their
own, a fundamental deficiency in Holland's view: ``[Expert] systems
are brittle in the sense that they respond appropriately only in
narrow domains, requiring substantial human intervention to compensate
for even slight shifts in domain'' \cite{Holland1986}. In
contrast, Holland proposed that ``{\emph induction} is the basic, and
perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines.  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Waldrop1993} (p. 182).

The if-then rules, when activated, compete to post their results on a
shared ``message list'', modeling the system's short-term memory
(again inspired by Hebb's work).  The competition was based on an
bidding process, in which a rule's current strength affected its
ability to bid highly.  Successful rules were gradually strengthened
(and unsuccessful rules weakened) via a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules would pay back part
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.  The idea for the
bucket-brigade algorithm was inspired by the mechanisms of Hebbian
learning and by early work on learning and optimization in economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1961,Samuel1959}). \footnote{Of course similar
  credit-assignment methods have been used extensively in
  reinforcement learning, neural networks, and other areas of modern
  machine learning, but Holland was among the first to propose such
  methods.}

Finally, new rules for the population are discovered by a genetic
algorithm, using the strength of rules as a fitness function.

Classifier systems are fairly complex architectures, with learning
occurring at multiple time scales.  Nevertheless, such systems have
been usefully applied to areas as diverse as
poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock
market \cite{Palmer1994}.  Modern reinforcement learning methods have
largely taken over as more useful approaches to real-world decision
and control problems, but classifier systems can perhaps be thought of
as an essential ``stage-setting'' method that enabled the development
of later approaches.

John Holland's main interest in classifier systems was how the two
learning mechanisms could work together to create subsets of rules
that accurately model the environment, and thus are able to ``fire''
at appropriate times.  An important part of such internal models is
the notion of a {\emph default hierarchy}---a set of rules that
together form a kind of hierarchical decision structure, going from
most general to most specific.  For example, a general rule might
state that, ``If X is a bird, then X can fly'', whereas a more specific
one might state, ``If X is a bird with small wings and a large body,
then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
the environment, both rules might compete to post their conclusion,
but the more specific one would be favored if both of its conditions are met.

Holland envisioned that, with a complex environment and enough time
for learning, a classifier system might learn an intricate default
hierarchy of rules, and that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it has learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
Offspring rules, which begin life weaker than do their parents, can
win the competition and influence the system's behavior only when
there are not strong rules whose conditions are satisfied---in other
words when the system does not know what to do.  If their actions
help, they survive; if not they are soon replaced.  Thus, the
offspring do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{Holland1992}.  A detailed
discussion of some real-world applications of classifier systems is
given in \cite{Booker1989}.

\section{Internal Models and Default Hierarchies}

The concept of {\emph internal models} is central to Holland's theory
of adaptive systems.  He posits that all adaptive systems create and
use internal models to prosper in their environments, through
anticipation.  Models can be tacit and learned through evolutionary
time, as in the case of bacteria swimming up a chemical gradient, or
explicit and learned over a single lifespan, as in the case of
cognitive systems that incorporate experience into internal rule sets
through learning.

In \cite{Holland1989}, Holland and his co-authors discuss how such
internal models can be learned, without supervision, by combining
environmental inputs with stored knowledge.  A key idea is that a
model defines sets of equivalence relations over environmental states,
together with a set of transition rules, which are learned over time
based on environmental ``rewards'' (or ``punishments'').  Models that
form valid homomorphisms with the environment allow the system to make
accurate predictions.  In Holland's conception, the equivalence
classes are initially very general, say ``moving object'' and
``stationary object,'' and over time, through experience and learning,
they are specialized into more useful and precise subclasses, say
``insect'' and ``nest''.  As described above, the adaptive system
eventually learns a default hierarchy of rules involving both general
and specific classes.

Although the idea of default hierarchies is prevalent in many
knowledge representation systems (LIST A FEW?), Holland made two key
contributions.  The first was his emphasis of homomorphisms as a formal way to
evaluate model validity.  This idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
%Today, even though computational modeling is a major
%application area of computing, essential to most natural sciences and
%engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of for characterizing when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update the model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism---classifier systems---by which a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment.  The key learning elements of this method, the
bucket-brigade algorithm, combined with a genetic algorithm, as
described above, presaged many of the ideas in modern reinforcement
learning.

>>>>

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic work
on integrating genetics with Darwinian selection \cite{Fisher1930}.
As Holland further read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain changing environments and use it in
order to improve their performance with respect to those environments.
Moreover, Holland recognized that in systems with {\em adaptive
  agents} of this kind, there is never a state of {\emph equilibrium}
or a final \emph{optimum} configuration.  Due to the environment's
perpetual novelty,
%SF thinks Holland did not invent this term (as Holland termed it), 
adaptation is a continual 
open-ended process.  Holland's focus on open-ended,
non-equilibrium dynamics was in stark contrast with the mainstream
approach (at the time) in all these fields---the belief that ``solving
for'' stable equilibrium dynamics was the scientific goal.  Holland's
contrary view was that a system in stable equilibrium is essentially
\emph{dead}.

% Text that STEPH found in our NIPS paper on the IGA
% First, at least theoretically the GA is fast because of implicit
% parallelism (Holland, 1975/1992): each string in the population is an
% instance of many different schemas, and if the population is large
% enough and is initially chosen at random, a large number of different
% schemas-many more than the number of strings in the population-are
% being sampled in parallel.  This should result in a quick search for
% short, low-order schemas that confer high fitness.  Second,
% fitness-proportionate reproduction under the GA should conserve
% instances of such schemas.  Third, a high crossover rate should
% quickly combine instances oflow-order schemas on different strings to
% create instances of longer schemas that confer even higher fitness.

Underlying Holland's theory of adaptation are three core ideas:
Population-based search, building blocks and recombination, and the
exploration-exploitation tradeoff.

\begin{itemize}
\item{\bf Populations and implicit parallelism:} A population of
  individuals can be thought of as a set of samples taken from a much larger
  search space.  Importantly, the populations are competitive, and
  the evolutionary dynamic biases the set of samples
  over time towards high-fitness regions of the search space.

%The population can be
%  thought of as a set of samples in a much larger search space, where
%  the processes of evolution
%  population can be thought of as a set of sa in which statistics 
%  can be leveraged to direct population dynamics.  (More on this below.)

\item{\bf Building Blocks and recombination:} High-fitness individuals
  are discovered in stages, first by finding high-fitness partial
  solutions (the building blocks) through random sampling, and over
  time assembling the partial solutions into more complete solutions
  with even higher fitness.  For example, when searching the space of
  all $n$-bit binary strings, almost any random sample would contain
  examples of strings whose first bit is set to 1, and also many
  samples of strings whose second bit is set to 1, but there would be
  many fewer samples that have both of the first two bits set to
  1.  More compelling examples include the \emph{Krebs cycle}, a core
  cellular metabolic pathway, a group of neurons that responds to an
  oriented edge in vision, or the ``swap'' operator in a sorting
  algorithm.

 To succeed, such a search requires first that
  high-fitness subparts be preserved once found, and second, that
  there be an efficient mechanism for combining promising building
  blocks into higher-order building blocks.  Holland proposed using an
  analog of Darwinian selection for the first and the idea of crossing
  over in genetics for the second.  Holland referred to the building
  blocks as \emph{schemas,} which correspond mathematically to hyperplanes in the
  search space.

%In a population undergoing adaptation,
%  individuals can be decomposed into {\emph building blocks}---sets of
% traits that are the evolutionary ``atoms'' of an individual's
%  fitness or performance.

\item{\bf Exploitation Versus Exploration:} Successful adaptation
  requires maintaining a balance between {\emph exploitation}, in which
  tried-and-true building blocks propagate in a population, and {\emph
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  
%A system that focuses too much on exploitation
%  risks never finding better individuals or not being able to adapt to
%  changing environments.  Conversely, too much focus on exploration
%  means that the system won't leverage the successful building blocks
%  that have already been discovered.  %There is some optimal balance
%  between these two modes of processing.
\end{itemize}

@@

Inspired by Bellman \cite{Bellman1961} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme {\emph
  exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme {\emph
  exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?
In \cite{Holland1973,Holland1975} Holland derived
an equation for such a strategy and argued that the optimal strategy
allocated trials to the observed best arm at a slightly higher rate than
exponential.  
% SF used 'argue' because some claim that his derivation was wrong.
He then extended this 
% Let $A$ denote the arm currently
%observed to have the higher payoff probability, and $B$ denote the
%other arm.  Holland showed that the optimal strategy (the one that
%yields maximal payoff, or minimal loss) is for the number of trials
%allocated to $A$ to grow slightly faster than an exponential function
%of the number of trials allocated to $B$.  Holland then showed that this
result to the {\emph multi-armed bandit} case.

In a population undergoing adaptation, building blocks consitute the arms of a multi-armed bandit.  Each
building block can be said to have a probability of ``payoff'' (i.e.,
contribution to a given individual's fitness).  Evaluating an
individual in an environment is like ``pulling the arms'' on a
multi-armed bandit, where the ``arms'' correspond to each of the
building blocks making up that individual.  Thus, in the process of
assigning a fitness to each individual in a population, adaptive
evolution can be seen as implicitly sampling the many building blocks
making up that collection of individuals.  The average fitness over
many individuals containing a particular building block gives an
estimate of that building block's payoff probability.

% The question of how to balance exploitation and exploration---how to
% optimally allocate trials to different arms based on their observed
% payoff---now becomes the question of how to optimally sample in the
% vast space of possible building blocks, based on their observed
% average fitness.  Of course adaptive evolution deals in populations of
% individuals, not building blocks.  There is no explicit mechanism for
% keeping statistics on building blocks' ``observed average fitness''.
% However, Holland's central idea here is that an approximation to
% optimal building-block sampling does indeed occur, as an emergent
% property of the population dynamics.

Holland argued that an approximation to
optimal building-block sampling actually occurs, as an emergent
property of the population dynamics defined by a
% To show this mathematically, he defined an idealization of
%adaptive evolution: an algorithm he called a {\emph reproductive
% plan}.
a population-based stochastic process
operating on bit strings, involving the ``genetic'' operators of
 reproduction, fitness-based selection, crossover, mutation, and
 inversion.  
%Building blocks are formalized as {\emph
%   schemata}---patterns of bits within a string. (Here are two examples
% of schemas: ``all strings beginning with the bits 10'' or ``all
% strings that start with the pattern 1*1'', where * can be replaced by
% either 0 or 1).

% Holland's most important result was the following: He proved
% mathematically that his reproductive plan results---implicitly---in a
% near-optimal allocation of trials to schemata, thus optimizing the
% exploitation versus exploration balance.  Holland noted that the
% explicit act of assigning fitnesses to the individuals in a population
% was actually implicitly sampling a much larger collection of building
% blocks.  Holland termed this {\emph implicit parallelism}; the
% parallel leveraging of statistics from implicit sampling is a main
% strength of his population-based approaches to search.

It was this formalization
% and resulting theorems that led to the
that led to the invention of genetic algorithms, which featured stochastic
population-based search and recombination as a critical
%{\emph crossover} as a critical
operation that allowed successful building blocks to be recombined and
tested in new contexts.  What made GAs unique
among other evolution-inspired algorithms at the time were the
mathematical foundations described above, the emphasis on populartions
and recombination as a central mechanism,
and a focus (at least in Holland's mind) on continual adaptation to
non-stationary environments rather than optimization to static
environments.

The framework Holland developed in \cite{Holland1975} was more general
than the genetic algorithm; its aim was an interdisciplinary theory of
adaptation, one that would inform biology, say, as much as computer
science \cite{Christiansen1998}.  The later, successful application of
genetic algorithms to real-world optimization and learning tasks was,
for Holland, just icing on the cake.

\section{Internal Models and Perpetual Novelty}

%One system of special interest to Holland was the human mind, and in
%particular, its capacity to learn via induction.

Internal models are central to Holland's theory of adaptive
systems.  He posits that all adaptive systems create and use internal
models to prosper in continually
changing environments---\emph{perpetual novelty}. Models can be tacit and learned
  through evolutionary time, as in the case of bacteria swimming up a
  chemical gradient, or explicit and learned over a single lifespan,
  as in the case of cognitive systems that incorporate experience into
  internal representations through learning.  In
  Holland's view, the key activity of an adaptive agent involves
  building and refining these data-driven models
  of the environment.
%using them to make predictions and take actions,
%  and occasionally receiving positive or negative feedbacks
%  (intermittent rewards).

In his second book, \emph{Induction} \cite{Holland1989}, Holland and
his co-authors tackled the question of how cognitive agents can
possibly learn and profit from
internal models, without supervision by combining
environmental inputs and rewards with stored knowledge.   In their
framework, a 
model defines a set of equivalence relations over environmental states,
together with a set of transition rules, which are learned over time
based on environmental rewards (or punishments).  Models that
form valid homomorphisms with the environment allow the system to make
accurate predictions.  In Holland's conception, the equivalence
classes are initially very general, say ``moving object'' and
``stationary object,'' and over time, through experience and learning,
 are specialized into more useful and precise subclasses, say
``insect'' and ``nest.''   Over time, the adaptive system
builds up a default hierarchy of rules covering general
cases and refinements for specific classes.

% For example, a general rule might
% state that, ``If X is a bird, then X can fly'', whereas a more specific
% one might state, ``If X is a bird with small wings and a large body,
% then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
% the environment, both rules might compete to post their conclusion,
% but the more specific one would be favored if both of its conditions are met.

% Maybe this should be moved to section on Complexity
Although the idea of default hierarchies was prevalent in 
knowledge representation systems of the era, Holland made two key
contributions.  The first was his emphasis on homomorphisms as a formal way to
evaluate model validity, and  idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
% %Today, even though computational modeling is a major
% %application area of computing, essential to most natural sciences and
% %engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of to characterize when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update the model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1977,Holland1986}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning, including unsupervised learning, non-Markovian learning, AND
ONE OTHER?.

% While genetic algorithms are nowadays most often associated with models of
% biological evolution, in Holland's view the processes of adaptation he
% was studying manifested themselves in many different complex systems.

% In the 1970s and 80s Holland formulated and extensively explored a
% particular model of decision-making, action, and inductive learning,
% called the {\emph classifier system} (e.g.,)
% .  In a classifier system, a population
% of ``if-then'' rules interacts with an environment, and over time is
% able to learn to improve its performance via both {\emph
%   credit-assignment} and {\emph rule-discovery} algorithms.

% Like genetic algorithms, classifier systems can be viewed both as 
% models of adaptation and as artificial-intelligence (AI) methods.  At
% the time Holland was developing classifier systems, the field of AI
% was focused on expert systems, which typically did not learn on their
% own, a fundamental deficiency in Holland's view: ``[Expert] systems
% are brittle in the sense that they respond appropriately only in
% narrow domains, requiring substantial human intervention to compensate
% for even slight shifts in domain'' \cite{Holland1986}. In
% contrast, Holland proposed that ``{\emph induction} is the basic, and
% perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1961,Samuel1959}).  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Waldrop1993} (p. 182).  
The if-then rules, when activated, compete to post their results on a
shared ``message list,'' modeling the system's short-term memory
(again inspired by Hebb's work and AI blackboard systems of the day).
Unlike the AI systems, however new rules were generated automatically,
in a trial-and-error fashion using a
genetic algorithm.

Successful rules were
strengthened over time if their predictions led to positive rewards
from the environment 
(and weakened otherwise).  
through a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules transferred some
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.   [ONE SENTENCE HERE
ABOUT SAMUEL'S CHECKER PLAYING PROGRAM.]
%, e.g., finding a target, or avoiding a predator,

%Finally, new rules for the population are discovered by a genetic
%algorithm, using the strength of rules as a fitness function.
%Classifier systems are fairly complex architectures, with learning
%occurring at multiple time scales.  
Although Holland proposed classifier systems as an executable theory
of inductive processes in cognition, other researchers took the
computational system more seriously, applying it to areas as diverse
as poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock market
\cite{Palmer1994}.  See Ref.~\cite{Booker1989} for more details about
practical applications of classifier systems.  Today, other
reinforcement learning methods are more popular for real-world
decision and control problems, but classifier systems can perhaps be
thought of as an essential ``stage-setting'' method that enabled the
development of later approaches.

% \footnote{Of course similar
%   credit-assignment methods have been used extensively in
%   reinforcement learning, neural networks, and other areas of modern
%   machine learning, but Holland was among the first to propose such
%   methods.}

Holland was primarily interested in how the two
learning mechanisms (discovery of new rules and apportioning credit to
existing rules) could work together to create 
%subsets of rules
%that accurately model the environment, and thus are able to ``fire''
%at appropriate times.  
%Holland envisioned that, with a complex environment and enough time
%for learning, a classifier system might learn an intricate default
useful default hierarchies of rules.  He emphasized that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it had learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
%Offspring rules, which begin life weaker than do their parents, can
%win the competition and influence the system's behavior only when
%there are not strong rules whose conditions are satisfied---in other
%words when the system does not know what to do.  If their actions
%help, they survive; if not they are soon replaced.  
%Thus,
... new rules do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{Holland1992}.
